 

### 2021-02-03 vertical-thermal-review.Rmd
organization/ writing
#41 , #33

## Introduction

@@

**Here we review how the biophysical environment and biological form and function vary across the vertical canopy gradient in forests.** 
We focus on five key themes (Fig. `r fig_schematic`):
(1) the biophysical environment;
(2) the leaf traits that most strongly influence leaf temperature and metabolism;
(3) resulting leaf temperature ($T_{leaf}$) and
(4) foliar metabolism, including stomatal conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
(5) resultant size-structuring of whole-tree and ecosystem function.
*We then consider the implications for modeling /  climate change responses.*

@@

@@

@@

Particularly in dense forests, the upper, outer canopy leaves of taller trees are exposed to dramatically different biophysical conditions than leaves in the canopy interior or understory (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`). 
Accordingly, there is substantive variation across the gradient in leaf traits (this section), which in turn shape leaf temperatures and metabolism (following sections).

@@

##### *(lifespan)*
*Leaf turnover tends to be greater in the sun, so of your standing stock of leaves (at least in evergreen species) sun leaves should be younger than shade leaves. This is bound to affect metabolism, and may affect other aspects of their performance.*

@@

**Many of the biophysical and trait variables reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig. `r fig_leaf_T`).
Leaf traits are influenced by biophysical environment in order to maintain metabolic optima so that the amount of carbon assimilated by a leaf throughout its life span is greater than the total carbon invested in the leaf [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020] 
$T_{leaf}$  is largely a function of radiation input versus heat lost to the environment, and typically increases [with tree height] as light availability and total radiation increases [Fig. `r fig_NEON_vertical`; @michaletzPlantThermoregulationEnergetics2015].

Greater upper canopy wind speeds (Fig. `r fig_NEON_vertical`) result in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with wind velocity [@daudetWindSpeedLeaf1999a]. 
Large $g_{a}$ to heat transfer, in turn, results in a smaller difference between $T_{air}$ and $T_{leaf}$, which is more typical of small leaves in the upper canopy, allowing them to dissipate heat more efficiently than larger leaves [@bauerleSeparatingFoliarPhysiology2011]. 
In low wind conditions, because $g_{a}$ to heat transfer is comparatively small, $T_{leaf}$ in the upper canopy can become substantially elevated in relation to $T_{air}$ under high radiation [@martin_boundary_1999; @leighInfluenceLeafSize2017]. 
Additionally, the thickness of the air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006] this tends to be greater in the understory than canopy? [ref]

In addition, leaf size, which itself varies with height (Table `r table_leaf_traits`), impacts leaf temperature. While small leaves remain within a few degrees of $T_{air}$, regardless of stomatal conductance, wider leaves can have temperatures deviating more from air temperature [@fauset_differences_2018]. Large leaves (with greater leaf width) can be significantly cooler than $T_{air}$ under low radiation with stomata open [@leighInfluenceLeafSize2017], and significantly hotter under high radiation with stomata closed.

Leaves with open stomata and high radiation loads maintain $T_{leaf}$ similar to $T_{air}$, with coolest leaves at intermediate sizes (~10mm). Shaded understory leaves should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance. However, counteracting this, lower wind speed in understory would reduce latent heat loss. Under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance. However, increase in air temperature typically also increases VPD, decrease in stomatal conductance during midday lead to a decrease in leaf CO2, and consequently increasing leaf temperature in upper canopy [@fauset_differences_2018]. Thus, their $T_{leaf}$ can be highly elevated above $T_{air}$ when stomatal conductance is limiting.

![**Figure `r fig_leaf_T`. Theoretical expectations for variation in the elevation of $T_{leaf}$ above $T_{air}$ in response to stomatal conductance ($g_s$), wind speed, and shortwave radiation.** Schematic produced using the tealeaves R package of @muir_tealeaves_2019. Current placeholder uses typical conditions for Harvard Forest, but model output deviates from real data (directions right, but Tleaf-Tair too high). [ISSUE #6](2020_12_draft).](leaf_energy_balance/2020_12_draft.png)

### In-situ observations
**Vertical profiles in microclimate (Fig. `r fig_NEON_vertical`) and leaf traits (Table `r table_leaf_traits`) together shape in-situ leaf temperatures according to the principles outlined above.** 
The net effect....

In a tropical wet and moist forest of Puerto Rico and temperate forests in Michigan, greater light input in exposed upper canopy leaves has been observed to increase leaf temperatures, sometimes 1-7C above ambient air temperature, making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@mauTemperateTropicalForest2018a]

-  Michaletz et al have done some interesting work on leaf thermoregulation in relation to various leaf traits. I don’t think there is sun vs shade tests, but we could look at how this might translate, or how some of the traits he has calculated might differ between sun and shade leaves. For example, he has this thermal time constant parameter that is calculated from leaf mass per area, leaf dry matter content, and some environmental parameters. E.g.: DOI: 10.1038/nplants.2016.129, http://dx.doi.org/10.1016/j.tree.2015.09.006

**When vertical gradients in the biophysical environment and leaf traits combine to shape leaf energy budgets, under some conditions $T_{leaf}$ tracks  $T_{air}$ and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).**
[Air temperature is not always a proxy for measuring leaf temperature [@michaletzPlantThermoregulationEnergetics2015].]
In a broadleaf deciduous forest, nighttime $T_{leaf}$'s were similar in the canopy and understory	(Bolstad et al. 1999). 
Similarly, during the dry season in a tropical moist forest in Panama... (Rey-Sánchez et al. 2016).

**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.**
-Large leaf to air temperature differences and high leaf temperatures were found in upper canopy of Atlantic, Brazil forests due to high solar radiation loads. However, leaf temperatures themselves were moderated by air temperature. [@fauset_differences_2018].
- In three forest types-- temperate, tropical wet and tropical moist, $T_{leaf max}$ exceeded temperature optimum for photosynthesis $T_{opt}$ in upper canopies of all sites, however $T_{leaf max}$ exceeded $T_{opt}$ at different heights and temperatures in all forests; with no significant relationship between $T_{opt}$ and height. Additionally, With increasing height, $T_{leaf} - T_{air}$ difference increased in the upper canopies of all forests, exhibiting greatest $T_{leaf} - T_{air}$ in temperate deciduous forest compared to tropical moist and wet.

-	(Slot et al. 2019) and refs therein
-	(Fauset et al. 2018)
Sun leaves can also be  cooled relative to $T_{air}$ more than shade leaves (Rey-Sánchez et al. 2016). There's some hint of this in Bolstad et al. 1999.

*(Because hydraulics are key to leaf T, we should include some of that here--e.g., water on leaves - evaporative cooling - and leaf water content)*

@@

### Stomatal conductance
Photosynthesis can be largely constrained by stomatal conductance. 
-"Sun exposed outer canopy leaves in tropical trees can experience temperatures up 48 C during diurnal cycle, midday stomtal closure reduces transpitrational cooling" thus elevating leaf temperature greater than air temperature (Slot et al. 2019)
- Max stomatal conductance - increases with canopy height (Kenzo et al. 2015)
Max transpiration rate - increases with canopy height (Kenzo et al. 2015).

@@

**Photosynthesis is generally higher in sun leaves-- a fact that is well-established (Niinemets 2007 ; Kenzo et al. 2015 ; Slot et al. 2019; Chen et al. 2020) and observed in numerous field studies.**
This is primarily driven by the greater light available to sun leaves.

In response to the greater light availability, sun leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions, including higher concentration of Rubsico, and multiple layers of palisade parenchyma. Light saturation level differs between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy)
The vertical gradient in photosythetic rates is also influenced by available concentration of N and P. (and water availability)
.[thoughts: plasticity vs change in tree species composition through the canopy; differences in variability of microenvironments through the profile - highest lower down in canopy?; life history / investment strategies and hence, leaf life spans at leaves at the top vs bottom of the canopy]-- M N. Smith [agreed. Again the importance of differentiating between within-species and among species trait variation. Also; steepness of the vertical gradient in light and VPD. In very tall trees a gradient in water potential might cause a gradient in stomatal limitations of net photosynthesis]

Photosynthesis increases with increasing irradiance, and saturates at quantum flux density of 400 $m^{-2}$ $s^{-1}$ to 700nm $m^{-2}$ $s^{-1}$ or greater in upper canopy. Photosynthesis has a peaked response to temperature, where photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al. 2017; Slot & Winter 2017). Beyond the optimum, photosynthesis decreases as a result of stomatal closure (e.g. Slot & Winter 2017; Smith et al. 2020 [temperate/boreal refs needed) and eventually due to biochemical constraints (refs. e.g. Sage & Kubien 2007; Varhammer et al. 2015 ).

@@

**While sun and shade leaves differ dramatically in $A_{max}$, the temperature sensitivity of photosynthesis does not appear to differ substantially between sun and shade leaves.**
Biophysically, we might expect that sun leaves should tend to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig. 14.5; *revisit this!*) (Campbell and Norman 1998) as observed by earlier studies, suggesting that the interaction between light gradient and temperature response along vertical canopy profile implied a spectrum of temperature responses of $T_{opt}$,$V_{max}$ and $J_{max}$ along the gradient [@niinemetsWithincanopyVariationRate2004a; Harley and Baldocchi 1995, Friend, 2001].
However, this does not appear to be the case based on the limited number of field studies that have compared temperature sensitivity of photosynthesis in sun and shade leaves.
For 3 species in Panama, @hernandezSimilarTemperatureDependence2020 and @slotPhotosyntheticHeatTolerance2019 found that the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant. Similar results were found for $V_{max}$ and $J_{max}$ [@slotPhotosyntheticHeatTolerance2019, Hernandez et al.2020]
Mau et al. found no trend in the optimum temperature for net photosynthesis along a height gradient in Puerto Rico, and no significant trend in temperate trees [@mau_temperate_2018].

**Leaf thermal tolerance (Tcrit/ T50)**
- define.
Typical values ~45-50$^\circ$C, varying somewhat 
across latitude/climate (O’sullivan et al. 2017),
with elevation [@feeley_thermal_2020;  Slot et al unpublished data],
with leaf traits [Sastry et al. 2018, @zhang_photosynthetic_2012].
Perez and Feeley et al. 2020 found that across species sun leaves with higher photosynthetic heat tolerance experienced higher maximum temperatures and had narrower thermal safety margins. These species may thus be at a greater risk of thermal damage.  
[- Elsa: also see: 
- Perez, T.M. and Feeley, K.J., 2020. Weak phylogenetic and climatic signals in plant heat tolerance. Journal of Biogeography.]

@@

## Ecological implications

**Differences across forest vertical gradients in biophysical conditions, plant traits, and metabolism and its thermal responses scale up to affect whole-tree ecology in several ways.** 
Vertical gradients in the biophysical environment shape which plant strategies, or sets of traits, are competitive in understory versus canopy conditions.

@@

**Vertical gradients affect tree growth and survival, and how these respond to climatic variation.**
There are numerous demographic differences between understory and canopy trees, which have been linked to both the physical environment and plant traits (e.g., REFS), and we will not attempt to review those here.
More interesting in this context is how these shape differences in the climatic sensitivity of growth and mortality.
Many observations of larger trees suffering more during drought (Bennett et al. 2015).
One likely mechanism behind this pattern is that the crowns of larger trees exist higher in the vertical profile and therefore are exposed to higher evaporative demand and solar radiation [@mcgregor_tree_nodate].
Although drought is primarily a hydraulic problem, lack of water --> lower gs --> higher leaf T, so leaves face tradeoffs of water loss vs potentially damaging leafT.

In northeastern US mesic temperate forests, tree-ring analyis [__time-scale] showed that the upper canopy was observed to be more sensitive to $VPD$, whereas the understory was insensitive [@rollinsonClimateSensitivityUnderstory, 2020]

**There is also recently emerging evidence that understory trees tend to exhibit greater growth declines with $T_{air}$ under mesic conditions.** 
In eight forests across the northeast United States, growth rates of understory trees declined sharply at higher $T_{air}$ [Fig `r fig_treerings` **???**; @rollinsonClimateSensitivityUnderstory].
This pattern is initially counter-intuitive in that sun-exposed leaves can experience greater elevation of $T_{leaf}$ over $T_{air}$ than shade leaves (Fig. `r fig_leaf_T` **???**). 
However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig. `r fig_leaf_T`).
It is also possible that competition is accelerated under warmer temperatures.

@@

Related to something in the NSC paragraph:
In this section we could perhaps also consider differences in biomass distribution between canopy trees an understory plants, and what that may mean for performance at high temperature. What I'm thinking is the temperature sensitivity of respiration that applies to leaves, stems, and roots, and even though thermal acclimation of leaf respiration is common, much less is known about acclimation of stem and root respiration, and incomplete acclimation still increases the respiration load at higher temperature. It is possible that an increased respiration load is more disadvantageous for understory plants because 1), I suspect that a greater fraction of their total biomass is life biomass that respires (as opposed dead heartwood in the stems of big trees), and 2), because they operate closer to their light compensation point, so if they lose more carbon in respiration, they would need  to make up for lost carbon in photosynthesis, but might lack sunlight to do so

@@

Microclimate warming has implications for forest regeneration [@vonarxSpatiotemporalEffectsForest2012a]. Temperature driven photosynthesis limitation is reducing tree growth rates, accompanying a global trend towards more frequently disturbed forests dominanted by younger trees [@mcdowellPervasiveShiftsForest2020]. Reductions in canopy cover can lead to non-linear threshold responses, causing dramatic shifts from one forest state to another (e.g., transition from forest to savanna-like vegetation in tropical forest regions through "savannization"), with energy balance impacts and implications for forest-atmosphere interactions [@starkReframingTropicalSavannization2020]. Each of these changes have potential feedbacks to climate change.

### Ecosystem Scale Implications

Leaf-scale properties and processes aggregate non-linearly  in vertical and horizontal space, influencing stand-level ecosystem processes, and in turn biosphere-atmosphere exchanges of water, carbon, energy, and trace gases. As described above, variation in temperature across vertical gradients can interact with local light conditions and other vertically varying meteorological conditions to affect species composition (e.g., see @nakamuraForestsTheirCanopies2017), leaf traits, and photosynthesis . Linked to this local biophysical variation is the partitioning of evapotranspiration (ET) into soil evaporation, canopy evaporation, and transpiration, which occur across varying spatiotemporal scales depending on light availability, water stress, stomatal physiology, and turbulent transfer (Lawrence et al. 2007, Staudt et al. 2011). Vertical microclimate gradients interact with vertical variation in foliage density, influencing not only the location of evapotranspiration in the canopy, but also differences in rates and net gross primary productivity (GPP, Banerjee & Linn 2018). It has been demonstrated that net carbon assimilation and transpiration can vary substantially for forest canopies with the same LAI but different leaf area density distributions (e.g., Katul et al 2004 , Banerjee & Linn 2018). Further, although transpiration tends to be the dominant component of ET, partitioning is more complex in open canopies where soil evaporation can become more dominant (Baldocchi et al. 2004, Ma et al. 2020), likely affecting vertical thermal gradients. In other words, it matters whether foliage is dense or sparse, and whether it is concentrated higher in the canopy, lower in the canopy, or evenly distributed throughout.

Still, partitioning ET into its constituent components, let alone vertically partitioning transpiration rates in the canopy, remains a challenge. Eddy covariance techniques, sap flow measurements, and stable isotope analysis offer useful tools, although errors can be large (Williams et al 2004, Herbst et al. 1996). To date, empirical analyses of ET vertical partitioning and interactions with temperature and vegetation structure have be concentrated in midlatitude ecosystems (e.g. Haverd et al 2009, 2011, Staudt et al. 2011, Ringgaard et al. 2014) *(KAT: but see Kunert et al. 2017 -not sure if this is equivalent)*. With respect to changing temperatures, the net effect of leaf- and plant-level thermal responses remains an active area of research, including both how vertical variation in thermal sensitivity influences forest ecosystem dynamics and biosphere-atmosphere interactions.

@@

[Elsa: do we want a section about research opportunities using remote sensing or is that outside the scope of this review?] 
An increasing availability of airborne and spaceborne LiDAR remote sensing data offers a promising opportunity for mapping vertical thermal gradients and vegetation structure. [Drone based FLIR cams, tower thermal cams (Pau et al 2018), ECOSTRESS]. To evaluate large-scale ecosystem patterns, high resolution vertically resolved vegetation structure from LiDAR data can be analyzed in combination with collocated field measurements collected using towers (FluxNet REF), canopy cranes (@nakamuraForestsTheirCanopies2017), or by tree climbing (REF – e.g. Asner papers, Shenkin et al – Unding Jami’s work). Existing research efforts have already revealed incredibly heterogeneity in vegetation structure, diversity, traits, and properties both within and across ecosystems (REFS – Stovall, Schneider, …others).

## Implications

@@

>>>>

# Introduction

@@

**Here we review how the biophysical environment and biological form and function vary across the vertical canopy gradient in forests.** 
We focus on five key themes (Fig. `r fig_schematic`):
(1) the biophysical environment;
(3) leaf temperature ($T_{leaf}$);
(2) the leaf traits most strongly influence leaf temperature and metabolism;
(4) foliar metabolism, including stomatal conductance, photosynthesis, respiration, and volatile organic compound (VOC) production; and 
(5) size-structuring of whole-tree and ecosystem function.
*We then consider the implications for scaling across space and time, and understanding forest responses to global change.*

@@

# Review of vertical gradients

@@

**Many of the biophysical and trait variables reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019; @michaletzEnergeticCarbonEconomic2016] (Fig. `r fig_leaf_T`).
Leaf traits are influenced by biophysical environment in order to maintain metabolic optima so that the amount of carbon assimilated by a leaf throughout its life span is greater than the total carbon invested in the leaf [@michaletzPlantThermoregulationEnergetics2015; @perezPhotosyntheticHeatTolerances2020] 
The difference between $T_{leaf}$ and $T_{air}$ is determined as a function of radiation input versus heat lost to the environment [Fig. `r fig_NEON_vertical`; @michaletzPlantThermoregulationEnergetics2015].

![**Figure `r fig_leaf_T`. Theoretical expectations for variation in the elevation of $T_{leaf}$ above $T_{air}$ in response to (a) shortwave radiation, (b) wind speed, (c) stomatal conductance ($g_s$), and (d) leaf width.** Schematic produced using the tealeaves R package of @muir_tealeaves_2019. Current placeholder uses typical conditions for Harvard Forest, but model output deviates from real data (directions right, but Tleaf-Tair too high). [ISSUE #6](2020_12_draft).](leaf_energy_balance/2020_12_draft.png)

**(paragraph on radiation as a driver)**
Shaded understory leaves should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance.

Greater upper canopy wind speeds (Fig. `r fig_NEON_vertical`) result in higher boundary layer conductance ($g_{a}$) for canopy leaves, which increases linearly with wind velocity [@daudetWindSpeedLeaf1999a]. 
Large $g_{a}$ to heat transfer, in turn, results in a smaller difference between $T_{leaf}$ and $T_{air}$. 
In low wind conditions, because $g_{a}$ to heat transfer is comparatively small, $T_{leaf}$ in the upper canopy can become substantially elevated in relation to $T_{air}$ under high radiation [@martin_boundary_1999; @leighInfluenceLeafSize2017]. 
Additionally, the thickness of the air layer is proportional to boundary layer resistance at the surface of the leaf through which water vapor diffuses after leaving the stomata [@martin_boundary_1999; @robertsStomatalBoundaryLayerConductances1990a; @stokesBoundaryLayerConductance2006] this tends to be greater in the understory than canopy? [ref]

**(paragraph on hydraulics as a driver of leaf T:e.g., water on leaves - evaporative cooling - and leaf water content)**
Leaves with open stomata can maintain $T_{leaf}$ similar to $T_{air}$ at high radiation loads, *with coolest leaves at intermediate sizes (~10mm)*. 
Under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance. 
However, increase in air temperature typically also increases VPD, decrease in stomatal conductance during midday lead to a decrease in leaf CO~2~, and consequently increasing leaf temperature in upper canopy [@fauset_differences_2018]. 
Thus, their $T_{leaf}$ can be highly elevated above $T_{air}$ when $g_{s}$ or $g_{a}$ is low.

**In addition, leaf size and morphology, which vary dramatically across the vertical gradient (reviewed in following section), impact $g_{a}$ and therefore leaf temperature.** 
*Smaller leaves have higher $g_{a}$ to heat transfer, allowing them to dissipate heat more efficiently than larger leaves [@bauerleSeparatingFoliarPhysiology2011]. (verify this statement and reference)* 
While small leaves remain within a few degrees of $T_{air}$, regardless of stomatal conductance, wider leaves can have temperatures deviating more from air temperature [@fauset_differences_2018]. 
Large leaves (with greater leaf width) can be significantly cooler than $T_{air}$ under low radiation with stomata open [@leighInfluenceLeafSize2017], and significantly hotter under high radiation with stomata closed.

### In-situ observations

**Vertical profiles in microclimate (Fig. `r fig_NEON_vertical`), leaf traits (see following section), and *evapotranspiration* (*where is this discussed? and do we say anything about evaporation?*) together shape in-situ leaf temperatures according to the principles outlined above (Fig. `r fig_leaf_T`).** 
While the high-light environment of the upper canopy would tend to dramatically increase $T_{leaf}$ above $T_{air}$, particularly under low-wind conditions, strong vertical gradients in leaf morphology and sometimes *evapotranspiration* have a counteractive effect, such that vertical trends in $T_{leaf}$ are much less pronounced than would be expected based on the physical setting alone. 
Under some conditions $T_{leaf}$ tracks  $T_{air}$ and therefore differs little between understory and canopy (Fig. `r fig_NEON_vertical`).
For instance, during the dry season in a tropical moist forest in Panama... (Rey-Sánchez et al. 2016). Sun leaves can also be  cooled relative to $T_{air}$ more than shade leaves (Rey-Sánchez et al. 2016). 
At night when solar radiation is not a factor, differences in $T_{leaf}$ across the vertical gradient are very modest; for instance, @bolstad_foliar_1999 observed similar nighttime $T_{leaf}$'s in the canopy and understory	of a temperate broadleaf forest.

**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.**
-Large leaf to air temperature differences and high leaf temperatures were found in upper canopy of Atlantic, Brazil forests due to high solar radiation loads. However, leaf temperatures themselves were moderated by air temperature. [@fauset_differences_2018].
- In three forest types-- temperate, tropical wet and tropical moist, $T_{leaf max}$ exceeded temperature optimum for photosynthesis $T_{opt}$ in upper canopies of all sites, however $T_{leaf max}$ exceeded $T_{opt}$ at different heights and temperatures in all forests; with no significant relationship between $T_{opt}$ and height. Additionally, With increasing height, $T_{leaf} - T_{air}$ difference increased in the upper canopies of all forests, exhibiting greatest $T_{leaf} - T_{air}$ in temperate deciduous forest compared to tropical moist and wet.
- In a tropical wet and moist forests of Puerto Rico and temperate forests in Michigan, greater light input in exposed upper canopy leaves has been observed to increase leaf temperatures, sometimes 1-7C above ambient air temperature, making upper canopy leaves susceptible to heat stress, more than scattered light in lower shaded leaves [@mauTemperateTropicalForest2018a] 
-"Sun exposed outer canopy leaves in tropical trees can experience temperatures up 48 C during diurnal cycle, midday stomtal closure reduces transpitrational cooling" thus elevating leaf temperature greater than air temperature (Slot et al. 2019)

-	(Slot et al. 2019) and refs therein

@@

Particularly in dense forests, the upper, outer canopy leaves of taller trees are exposed to dramatically different biophysical conditions than leaves in the canopy interior or understory (Figs. `r fig_NEON_vertical`, `r SIfigs_NEON_all`).
Leaf traits are critical to maintaining desirable $T_{leaf}$ and, in turn, shaping leaf metabolism and carbon balance (following section).
It is unsuprising, then, that leaf traits vary dramatically across vertical gradients.

@@

##### *(lifespan)*
*Leaf turnover tends to be greater in the sun (in the tropics), so of your standing stock of leaves (at least in evergreen species) sun leaves should be younger than shade leaves. This is bound to affect metabolism, and may affect other aspects of their performance.*

@@

@@

### Stomatal conductance
Photosynthesis can be largely constrained by stomatal conductance.

- Max stomatal conductance - increases with canopy height (Kenzo et al. 2015)
Max transpiration rate - increases with canopy height (Kenzo et al. 2015).

@@

**Photosynthesis is generally higher in sun leaves-- a fact that is both theoretically expected and observed in numerous field studies (Table 2; Niinemets 2007 ; Kenzo et al. 2015 ; Slot et al. 2019; Chen et al. 2020).**
This is primarily driven by the greater light available to sun leaves.

In response to the greater light availability, sun leaves have traits allowing greater photosynthetic rates at high light under sufficient nutrient and water conditions, including higher concentration of Rubsico, and multiple layers of palisade parenchyma. Light saturation level differs between leaves and canopies, with leaves saturating at lower light level than canopies (while a certain amount of light saturates the photosynthetic capacity of the leaf, increasing the amount of light in the canopy may increase photosynthesis in the shaded part of the canopy)
The vertical gradient in photosynthetic rates is also influenced by available concentration of N and P. (and water availability)
*[thoughts: plasticity vs change in tree species composition through the canopy; differences in variability of microenvironments through the profile - highest lower down in canopy?; life history / investment strategies and hence, leaf life spans at leaves at the top vs bottom of the canopy]-- M N. Smith [agreed. Again the importance of differentiating between within-species and among species trait variation. Also; steepness of the vertical gradient in light and VPD. In very tall trees a gradient in water potential might cause a gradient in stomatal limitations of net photosynthesis]*

Photosynthesis increases with increasing irradiance, and saturates at quantum flux density of 400 $m^{-2}$ $s^{-1}$ to 700nm $m^{-2}$ $s^{-1}$ or greater in upper canopy.

Photosynthesis has a peaked response to temperature, where photosynthesis is commonly found to be maximized at the prevalent ambient growing season conditions (Tan et al. 2017; Slot & Winter 2017). Beyond the optimum, photosynthesis decreases as a result of stomatal closure (e.g. Slot & Winter 2017; Smith et al. 2020 [temperate/boreal refs needed) and eventually due to biochemical constraints (refs. e.g. Sage & Kubien 2007; Varhammer et al. 2015 ).

**While sun and shade leaves differ dramatically in $A_{max}$, the temperature sensitivity of photosynthesis does not appear to differ substantially between sun and shade leaves.**
Biophysically, we might expect that sun leaves should tend to have a stronger temperature-dependence and higher temperature optima than shade leaves (Fig. 14.5 in Campbell and Norman 1998; *revisit this!*) *as observed by earlier studies, suggesting that the interaction between light gradient and temperature response along vertical canopy profile implied a spectrum of temperature responses of $T_{opt}$,$V_{max}$ and $J_{max}$ along the gradient [@niinemetsWithincanopyVariationRate2004a; Harley and Baldocchi 1995, Friend, 2001].* *(KAT: I'm confused by the second part of this sentence. Are these studies, or theoretical expectations?.)*
However, such a trend is not apparent based on the limited number of field studies that have compared temperature sensitivity of photosynthesis in sun and shade leaves.
For 3 species in Panama, the optimum temperature for sun leaves tended to be slightly higher than that of shade leaves, but differences were not significant [@slotPhotosyntheticHeatTolerance2019 ; @hernandezSimilarTemperatureDependence2020]. Similar results were found for $V_{max}$ and $J_{max}$ [@slotPhotosyntheticHeatTolerance2019, Hernandez et al.2020]
Similarly, Mau et al. found no trend in the optimum temperature for net photosynthesis along a height gradient in Puerto Rico, and no significant trend in temperate trees [@mau_temperate_2018].

*(KAT: let's cut the portion on T50 down to one or at most two par:)*

@@

**Leaf thermal tolerance (Tcrit/ T50)**
*(define)*
Typical values ~45-50$^\circ$C, varying somewhat 
across latitude/climate (O’sullivan et al. 2017),
with elevation [@feeley_thermal_2020;  Slot et al unpublished data],
with leaf traits [Sastry et al. 2018, @zhang_photosynthetic_2012].
Perez and Feeley et al. 2020 found that across species sun leaves with higher photosynthetic heat tolerance experienced higher maximum temperatures and had narrower thermal safety margins. These species may thus be at a greater risk of thermal damage.  
[- Elsa: also see: 
- Perez, T.M. and Feeley, K.J., 2020. Weak phylogenetic and climatic signals in plant heat tolerance. Journal of Biogeography.]

@@

## Size-structuring within the ecosystem

**Differences across forest vertical gradients in biophysical conditions, plant traits, and metabolism and its thermal responses scale up to affect whole-tree ecology and ecosystem ecology in several ways.** 
Vertical gradients in the biophysical environment shape which plant strategies, or sets of traits, are competitive in understory versus canopy conditions.

@@

**Vertical gradients affect tree recruitment, growth and survival, and how these respond to climatic variation.**
There are numerous demographic differences between understory and canopy trees, which have been linked to both the physical environment and plant traits (e.g., REFS), *and we will not review those here.* *(although maybe we should put one paragraph?)*

More interesting in this context is how these shape differences in the climatic sensitivity of growth and mortality.
Many observations of larger trees suffering more during drought (Bennett et al. 2015).
One likely mechanism behind this pattern is that the crowns of larger trees exist higher in the vertical profile and therefore are exposed to higher evaporative demand and solar radiation [@mcgregor_tree_nodate].
Although drought is primarily a hydraulic problem, lack of water --> lower gs --> higher leaf T, so leaves face tradeoffs of water loss vs potentially damaging leafT.

**There is also recently emerging evidence that understory trees tend to exhibit greater growth declines with $T_{air}$ under mesic conditions.** 
In eight forests across the northeast United States, growth rates of understory trees declined sharply at higher $T_{air}$ [Fig `r fig_treerings` **???**; @rollinsonClimateSensitivityUnderstory].
This pattern is initially counter-intuitive in that sun-exposed leaves can experience greater elevation of $T_{leaf}$ over $T_{air}$ than shade leaves (Fig. `r fig_leaf_T` **???**). 
However, given adequate moisture to sustain high $g_s$, canopy leaves exposed to more wind and lower humidity should be more effective at cooling when $T_{air}$ exceeds physiological optima (Fig. `r fig_leaf_T`).
It is also possible that higher temperatures reduce understorty tree growth in these forests through an indirect mechanism, such as accelerated competition under warmer temperatures.

@@

*In this section we could perhaps also consider differences in biomass distribution between canopy trees an understory plants, and what that may mean for performance at high temperature. What I'm thinking is the temperature sensitivity of respiration that applies to leaves, stems, and roots, and even though thermal acclimation of leaf respiration is common, much less is known about acclimation of stem and root respiration, and incomplete acclimation still increases the respiration load at higher temperature. It is possible that an increased respiration load is more disadvantageous for understory plants because 1), I suspect that a greater fraction of their total biomass is life biomass that respires (as opposed dead heartwood in the stems of big trees), and 2), because they operate closer to their light compensation point, so if they lose more carbon in respiration, they would need  to make up for lost carbon in photosynthesis, but might lack sunlight to do so*

@@

### C and water flux

@@

Leaf-scale properties and processes aggregate non-linearly  in vertical and horizontal space, influencing stand-level ecosystem processes, and in turn biosphere-atmosphere exchanges of water, carbon, energy, and trace gases. As described above, variation in temperature across vertical gradients can interact with local light conditions and other vertically varying meteorological conditions to affect species composition (e.g., see @nakamuraForestsTheirCanopies2017), leaf traits, and photosynthesis . Linked to this local biophysical variation is the partitioning of evapotranspiration ($ET$) into soil evaporation, canopy evaporation, and transpiration, which occur across varying spatiotemporal scales depending on light availability, water stress, stomatal physiology, and turbulent transfer (Lawrence et al. 2007, Staudt et al. 2011). Vertical microclimate gradients interact with vertical variation in foliage density, influencing not only the location of evapotranspiration in the canopy, but also differences in rates and net gross primary productivity ($GPP$, Banerjee & Linn 2018). It has been demonstrated that net carbon assimilation and transpiration can vary substantially for forest canopies with the same LAI but different leaf area density distributions (e.g., Katul et al 2004 , Banerjee & Linn 2018). Further, although transpiration tends to be the dominant component of ET, partitioning is more complex in open canopies where soil evaporation can become more dominant (Baldocchi et al. 2004, Ma et al. 2020), likely affecting vertical thermal gradients. In other words, it matters whether foliage is dense or sparse, and whether it is concentrated higher in the canopy, lower in the canopy, or evenly distributed throughout.

Still, partitioning $ET$ into its constituent components, let alone vertically partitioning transpiration rates in the canopy, remains a challenge. Eddy covariance techniques, sap flow measurements, and stable isotope analysis offer useful tools, although errors can be large (Williams et al 2004, Herbst et al. 1996). To date, empirical analyses of ET vertical partitioning and interactions with temperature and vegetation structure have be concentrated in midlatitude ecosystems (e.g. Haverd et al 2009, 2011, Staudt et al. 2011, Ringgaard et al. 2014) *(KAT: but see Kunert et al. 2017 -not sure if this is equivalent)*. With respect to changing temperatures, the net effect of leaf- and plant-level thermal responses remains an active area of research, including both how vertical variation in thermal sensitivity influences forest ecosystem dynamics and biosphere-atmosphere interactions.

# Implications

## Scaling across space and time

@@

### Remote sensing
